---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

In this section we read the data and do a bit of processing to simplify plotting.

```{r}
# Options for number formatting, load libraries
options(scipen=2, digits=2)
library(data.table)
library(ggplot2)
# Read the data, unz helps us read the zip-file directly
data <- read.csv(unz("activity.zip", "activity.csv"), colClasses=c("numeric", "Date", "integer"))
DT <- data.table(data)
# Create a time field (in hours) for plotting, since the interval is not suitable for this
# Since this updates in place, wrap in invisible() to suppress output
invisible(DT[, `:=`(ihour=as.numeric((interval%/%100)+(interval%%100)/60.0))])
# Support function for plotting time values (7.75 ->  07:45)
timeFormatter <- function(hours) {
    h <- hours%/%1
    m <- round(60*(hours%%1))
    sprintf("%02d:%02d", h, m)
}
```

## What is mean total number of steps taken per day?
```{r}
# Create a sum of steps by date
DS <- DT[!is.na(steps),.(stepsum=sum(steps)), by=date]
# Plot a histogram
hist(DS$stepsum, xlab="Number of steps", main="Histogram of total daily steps")
# Calculate mean and median
meansteps <- mean(DS[,stepsum])
mediansteps <- median(DS[,stepsum])
```

The mean numer of steps per day is `r meansteps` and the median `r mediansteps` steps.


## What is the average daily activity pattern?
```{r}
# Calculate averages for each interval
AD <- DT[, .(smean=mean(steps, na.rm=T)), .(interval,ihour)]

# plot
qplot(data=AD, x=ihour, y=smean, geom="line", xlab="Time of day",ylab="Mean number of steps") + scale_x_continuous(breaks=c(0,6,12,18,24), label=timeFormatter) + ggtitle("Average daily activity pattern")

# Find the top interval
top_interval <- AD[smean==(AD[, max(smean)]), interval]
thour = top_interval %/% 100
tmin = top_interval %% 100
```

The interval with the highest number of steps on average is the interval `r top_interval`. That is the interval starting at hour `r thour`, minute `r tmin`.


## Imputing missing values

Impute missing values by using the mean for that interval.

```{r}
# How many missing values are there?
missingvals <- DT[is.na(steps), .("Missing values"=.N)]
missingvals
# Create a new dataset where we'll impute missing values
DTI <- DT
# Add the mean values for each interval to the table
invisible(DTI[,`:=`(imean=mean(steps, na.rm=T)), by=interval])
# Use the mean values for the interval to impute missing values
invisible(DTI[is.na(steps),`:=`(steps=imean)])
# Create a sum of steps by date
DSI <- DTI[,.(stepsum=sum(steps)), date]
hist(DSI$stepsum, xlab="Number of steps", main="Histogram of total daily steps")
imeansteps <- mean(DSI[,stepsum])
imediansteps <- median(DSI[,stepsum])
```
There were `r missingvals` missing values before imputing.  
After imputing, the mean number of steps per day is `r imeansteps` and the median `r imediansteps` steps. Since we use the mean to impute the missing values, it is unchanged, but the median has changed, the old median was `r mediansteps`.


## Are there differences in activity patterns between weekdays and weekends?

To separate weekdays from weekends we add weekdays to the table, then apply a custom function to check if it is a weekday or weekend, storing the result as a factor.

```{r}
# Add weekday names to the table
invisible(DTI[,`:=`(wd=weekdays(strptime(date, format="%Y-%m-%d")))])
# Support function to map days to weekday/weekend
daytype <- function(dayname) {
    weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
    if(dayname %in% weekdays) {
        return(factor("weekday", c("weekday", "weekend")))
    }
    factor("weekend", c("weekday", "weekend"))
}
# Apply the function above to the weekday data
invisible(DTI[,we:=sapply(wd, daytype)])
# Calculate means for each interval on weekdays/weekends
WS <- DTI[, .(smean=mean(steps, na.rm=T)), by=.(interval, ihour, we)]
# Plot the result
qplot(data=WS, x=ihour, y=smean, facets=we~., geom="line", xlab="Time of day",ylab="Mean number of steps") + scale_x_continuous(breaks=c(0,6,12,18,24), label=timeFormatter) + ggtitle("Activity patterns weekday/weekend")
```

As can be seen in the plot, activity starts later on weekends but goes on for longer in the evening.
